---
title: "Primary tumour DE collate results"
author: "Kieran R Campbell"
output:
  html_document: 
    toc: true
    toc_float: true
  html_notebook: default
params:
  input_rds: '../../data/pdx_temp_de/v3/DE_results_pseudobulk_FALSE.rds'
  cellranger_version: v3
  pseudobulk: FALSE
  volcano_plot: volcano.png
  grid_plot: grid.png
  pathway_plot: pathway.png
  coregene_stats: coregene.stats
  coregene_csv: coregene.csv
  pathway_membership_plot: pathplot.png
---


```{r}
knitr::opts_chunk$set(echo = TRUE,
                      cache = FALSE)

suppressPackageStartupMessages({
  library(tidyverse)
  library(glue)
  library(ggrepel)
  library(org.Hs.eg.db)
  library(here)
  library(forcats)
})

select <- dplyr::select

pseudobulk <- as.logical(params$pseudobulk)

source(here("scripts/utils.R"))
```


Load in data



```{r}
results <- readRDS(params$input_rds)

df <- results$edger_results

df_pt <- results$camera_results
```


# Data summaries

Get the config we need:

```{r}
config <- get_config()

alpha <- config$global_significance
core_thresh <- config$core_geneset_fc
```


Number of cells: `r nrow(results$design)`

Total genes significant at 5% FDR:

```{r}
table(df$FDR < alpha)
```



```{r}
coregene_df <- filter(df, FDR< alpha, abs(logFC) > log2(core_thresh))

df <- mutate(df, is_coregene = ensembl_gene_id %in% coregene_df$ensembl_gene_id)

```

In total this gives `r nrow(coregene_df)` genes

Up - down regulation:

```{r}
table(sign(coregene_df$logFC))
```

# Write coregene df to file

```{r}
write_csv(coregene_df, params$coregene_csv)
```


# Volcano plots



```{r}
df_ann <- coregene_df %>% 
  top_n(20, abs(logFC))
```


```{r}
ggplot(df, aes(x = logFC, y = -log10(FDR))) +
  geom_point(aes(colour = is_coregene)) +
  geom_hline(yintercept = -log10(alpha), colour = 'darkred', linetype = 2) +
  cowplot::theme_cowplot(font_size = 11) +
  geom_label_repel(aes(label = gene_symbol), data = df_ann) +
  scale_colour_manual(values = c("TRUE"="black", "FALSE"="grey60"), guide = FALSE) +
  labs(x = expression(log[2]~"(fold change) 37C vs 6C"), y = expression(-log[10]~"(q-value)")) 
```

```{r}
saveRDS(last_plot(), params$volcano_plot)
```



# Core gene set plots

```{r}

symbol_na <- is.na(coregene_df$gene_symbol)

coregene_df$gene_symbol[symbol_na] <- coregene_df$ensembl_gene_id[symbol_na]
```


```{r}
coregene_df %>% 
  top_n(min(40, nrow(coregene_df)), abs(logFC)) %>% 
  ggplot(aes(x = forcats::fct_reorder(gene_symbol, logFC), y = logFC, fill = logFC)) +
  geom_bar(stat = 'identity') +
  scale_fill_gradient2(low = scales::muted("blue"), high = scales::muted("red"), guide = FALSE) +
  cowplot::theme_cowplot(font_size = 11) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
      legend.position = "top") +
  labs(x = "Gene", y = expression(log[2]~"(fold change) 37C vs 6C")) +
  # geom_hline(yintercept = log2(1.5), colour = 'grey50', linetype = 1, size = 1) +
  coord_flip()
```

```{r}
saveRDS(last_plot(), params$grid_plot)
```

# Pathway analysis



```{r}
filter(df_pt, FDR < alpha) %>% 
ggplot(aes(x = forcats::fct_reorder(gsub("HALLMARK_", "", pathway), -FDR), y = -log10(FDR))) +
  geom_bar(aes(fill = Direction), stat = 'identity') +
  coord_flip() +
  scale_fill_manual(values = c("Up"=scales::muted('red'),
                               "Down"=scales::muted('blue')), 
                    name = "Regulation at 37C") +
  cowplot::theme_cowplot(font_size = 11) +
  labs(x = "Pathway", y = expression(-log[10]~"(FDR)"), subtitle = "Significant pathways") +
  theme(legend.position = c(0.6, 0.1),
        legend.box.background = element_rect(colour = 'black', size = .2))
```



Let's read in the original pathways


```{r}
load(here("data/genesets/human_H_v5p2.rdata"))
go_gs <- Hs.H

entrezgene_symbol_map <- as.list(org.Hs.egSYMBOL)


map_ids <- function(entrezgenes) {
  x <- unlist(entrezgene_symbol_map[entrezgenes])
  names(x) <- NULL
  x
}
```

```{r}
go_gs_symbol <- lapply(go_gs, map_ids)

core_genes <- coregene_df$gene_symbol

n_core_genes <- lapply(go_gs_symbol, function(symbols) sum(core_genes %in% symbols))

n_df <- tibble(pathway = names(n_core_genes), 
               pathway_size = sapply(go_gs_symbol, length),
               n_genes = unlist(n_core_genes),
               pct_pathway_genes = n_genes / pathway_size * 100)
```

```{r}
print(n_df)
```


```{r}
filter(df_pt, FDR < 0.05) %>% 
  inner_join(n_df) %>% 
  mutate(pathway = forcats::fct_reorder(gsub("HALLMARK_", "", pathway), -FDR)) %>% 
  ggplot(aes(x = pathway, y = -log10(FDR))) +
  geom_segment(aes(y = 0, yend = -log10(FDR), x = pathway, xend = pathway), linetype = 2, colour = 'grey50') +
  geom_point(aes(size = pct_pathway_genes, fill = Direction), shape = 21) +
  coord_flip() +
  scale_fill_manual(values = c("Up"=scales::muted('red'),
                               "Down"=scales::muted('blue')), 
                    name = "Regulation at 37C") +
  scale_size(name = "% pathway genes\nin core geneset") +
  cowplot::theme_cowplot(font_size = 11) +
  labs(x = "Pathway", y = expression(-log[10]~"(FDR)"), subtitle = "Significant pathways") +
  theme(legend.position = c(0.5, 0.4),
        legend.box.background = element_rect(colour = 'black', size = .2))

```

```{r}
saveRDS(last_plot(), params$pathway_plot)
```


# Save various statistics on the core geneset

```{r}
n_genes_coreset <- nrow(coregene_df)
n_genes_coreset_up <- filter(coregene_df, logFC > 0) %>% nrow()
n_genes_coreset_down <- filter(coregene_df, logFC < 0) %>% nrow()

pct_genes_nfkb <- filter(n_df, pathway == "HALLMARK_TNFA_SIGNALING_VIA_NFKB") %>% .$pct_pathway_genes

df_stat <- frame_data(
  ~ description, ~ statistic,
  "n_genes_coreset", n_genes_coreset,
  "n_genes_coreset_up", n_genes_coreset_up,
  "n_genes_coreset_down", n_genes_coreset_down,
  "pct_genes_nfkb", pct_genes_nfkb
)

write_statistics(df_stat, file = params$coregene_stats)
```



# Pathway membership supplementary

```{r}
load(here("data/genesets/human_H_v5p2.rdata"))
go_gs <- Hs.H

entrezgene_ensembl_map <- as.list(org.Hs.egENSEMBL)

map_ids <- function(entrezgenes) {
  x <- unlist(entrezgene_ensembl_map[entrezgenes])
  names(x) <- NULL
  x
}
```

```{r}
go_gs_ensembl <- lapply(go_gs, map_ids)
names(go_gs_ensembl) <- sub("HALLMARK_", "", names(go_gs_ensembl))
```

```{r}
pathmem <- lapply(names(go_gs_ensembl), function(n) tibble(pathway = n, ensembl_gene_id = go_gs_ensembl[[n]])) %>% 
  bind_rows()
```

```{r}
pathmem <- inner_join(pathmem, coregene_df)

pathmem <- dplyr::count(pathmem, pathway, name='n_genes_in_pathway') %>% 
  inner_join(pathmem)

pathmem <- dplyr::count(pathmem, ensembl_gene_id, name='n_pathways_of_gene') %>% 
  inner_join(pathmem)

```

```{r}
top_n(coregene_df, 40, (logFC)) %>% 
  inner_join(pathmem) %>% 
  ggplot(aes(y = fct_reorder(gene_symbol, n_pathways_of_gene), 
                    x = fct_reorder(pathway, -n_genes_in_pathway))) +
  geom_tile() +
  labs(y = "Gene", x = "Pathway") +
  theme_paper() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

ggsave(params$pathway_membership_plot, width = 6, height = 9)
```


